Preventing runtime errors of Redis at compile time

### 2015-11-17 paper.tex
Section 3: Indexed Monad

%
% LaTeX template for prepartion of submissions to PLDI'16
%
% Requires temporary version of sigplanconf style file provided on
% PLDI'16 web site.
%
\documentclass[pldi]{sigplanconf-pldi16}
% \documentclass[pldi-cameraready]{sigplanconf-pldi16}

@@

% \usepackage[utf8]{inputenc}
% \usepackage[english]{babel}
\usepackage{minted}     % code highlighting
\usemintedstyle{tango}

@@

\title{Catching runtime errors of Redis at compile time}

@@

In this paper, we demonstrates how to prevent those runtime errors at compile
time, allowing users to write reliable database queries without runtime
overhead, by exploiting type-level programming techniques such as parametedzied
monad, type-level literals and type families in Haskell.

@@

\begin{minted}{text}
redis> SADD some-set a b c
(integer) 3
redis> SADD another-set a b
(integer) 2
redis> SINTER some-set another-set
1) "a"
2) "b"
\end{minted}

@@

\subsection{Hedis}
Redis can also be used in most programming languages, with 3rd party libraries
or packages that talk with Redis' TCP protocol. In Haskell, the most popular
package is \emph{Hedis}.

The previous example in Redis's client would look something like this with
\emph{Hedis} in Haskell.
\begin{minted}{haskell}
program :: Redis (Either Reply [ByteString])
program = do
    sadd "some-set" ["a", "b"]
    sadd "another-set" ["a", "b", "c"]
    sinter ["some-set", "another-set"]
\end{minted}

Function \mintinline{haskell}{sadd} takes a key and a list of values as
arguments, and returns an \mintinline{haskell}{Integer}, wrapped in
\mintinline{haskell}{Either} to indicate possible failures in the context
\mintinline{haskell}{Redis} of command execution.

\begin{minted}{haskell}
sadd :: ByteString      -- key
    -> [ByteString]     -- values
    -> Redis (Either Reply Integer)
\end{minted}

@@

\subsection{Semantics errors}

Although everything in Redis is binary string, they are treated diffenrently.
Redis supports many different kind of data structures, such as strings, hashes,
 lists, etc. These values are of different data types, and some commands only
 work on certain type of values.

@@

\begin{minted}{text}
redis> SET some-string foo
OK
redis> SADD some-string bar
(error) WRONGTYPE Operation against a key
 holding the wrong kind of value
\end{minted}

@@

\begin{minted}{text}
redis> SET some-string foo
OK
redis> INCR some-string
(error) ERR value is not an integer or out
 of range
\end{minted}

\paragraph{In Hedis} The same goes for Hedis, since it's only a simple wrapping on top of Redis's
protocol written in Haskell. No semantic checking is done, neither
dynamically nor statically.

\begin{minted}{haskell}
program :: Redis (Either Reply Integer)
program = do
    set "some-string" "foo"
    sadd "some-string" ["a"]
\end{minted}

Which yields the same error as in Redis's client.

\begin{minted}{haskell}
Left (Error "WRONGTYPE Operation against a
 key holding the wrong kind of value")
\end{minted}

@@

Haskell makes it easy to build and use Domain Specific Languages (DSLs).
Hedis can be regardred as one DSL that embeds in Haskell, that has variable
bindings (between keys and values), but with very little or none semantic
checkings.

We started out from making Hedis a dynamically checked embedded DSL, with a
runtime type checker that keeps track of everything. But then we found that,
things can be a lot easier, by laveraging the host language's type checker,
and make it statically checked!

@@

To summerize our contributions:

\begin{itemize}[noitemsep]
\item We model variable bindings of an embedded DSL with language extensions
like type-level literals and data kinds.
\item We provide (yet another) an example of how to encode effects and
constraints of an action in types, with indexed monad and other language
 extensions such as closed type-families and constraints kinds.
\item Edis, a package we built for programmers. This package helps programmers
to write more reliable Redis codes, and makes Redis polymorphic by converting
back and forth from values of any types and boring ByteStrings.
\end{itemize}

\section{Implementation}
\subsection{Model bindings with Type-level Dictionaries}
\paragraph{Type-level literals}
\subsection{Bookkeeping with Type families}
\subsection{Capture effects with Parameterised Monad}
\paragraph{Type-level literals}

>>>>

%
% LaTeX template for preparation of submissions to PLDI'16
%
% Requires temporary version of sigplanconf style file provided on
% PLDI'16 web site.
%
\documentclass[pldi]{sigplanconf-pldi16}
% \documentclass[pldi-cameraready]{sigplanconf-pldi16}

@@

% code highlighting
\usepackage{minted}
\usemintedstyle{tango}

% no red boxes on parser error:
\AtBeginEnvironment{minted}{%
  \renewcommand{\fcolorbox}[4][]{#4}}

% quotes
\usepackage{dirtytalk}

% for symbols
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{upquote}

% use arbitrary fonts
\usepackage{fontspec}
\setmonofont{Inconsolata}

@@

\title{Preventing runtime errors of Redis at compile time}

@@

In this paper, we demonstrates how to prevent those runtime errors at compile
time, allowing users to write more reliable database queries without runtime
overhead, by exploiting type-level programming techniques such as indexed
monad, type-level literals and closed type families in Haskell.

@@

\begin{minted}[xleftmargin=1em]{text}
redis> SADD some-set a b c
(integer) 3
redis> SADD another-set a b
(integer) 2
redis> SINTER some-set another-set
1) "a"
2) "b"
\end{minted}

@@

\subsection{Hedis}
Redis can also be used in most programming languages, with 3rd party libraries
or packages that talk with Redis' TCP protocol. In Haskell, the most popular
package is Hedis.

The previous example in Redis's client would look something like this with
Hedis in Haskell.
\begin{minted}[xleftmargin=1em]{haskell}
program :: Redis (Either Reply [ByteString])
program = do
    sadd "some-set" ["a", "b"]
    sadd "another-set" ["a", "b", "c"]
    sinter ["some-set", "another-set"]
\end{minted}

Function \mintinline{haskell}{sadd} takes a key and a list of values as
arguments, and returns an \mintinline{haskell}{Integer}, wrapped in
\mintinline{haskell}{Either} to indicate possible failures in the context
\mintinline{haskell}{Redis}\footnotemark of command execution.

\begin{minted}[xleftmargin=1em]{haskell}
sadd :: ByteString      -- key
    -> [ByteString]     -- values
    -> Redis (Either Reply Integer)
\end{minted}

@@

\footnotetext{Hedis provides another kind of context, \mintinline{haskell}{RedisTx}, for \emph{transactions}, united with \mintinline{haskell}{Redis} under the class of \mintinline{haskell}{RedisCtx}. We use \mintinline{haskell}{Redis} only here for brevity. }

\subsection{Problems}

\say{All binary strings are equal, but some binary strings are more equal than others.}

Although everything in Redis is essentially a binary string, they are treated
differently. Redis supports many different kind of data structures, such as
strings, hashes, lists, etc. These values are of different data types, and some
commands only work on certain types of them, much like how language C treats
a piece of data.

@@

\begin{minted}[xleftmargin=1em]{text}
redis> SET some-string foo
OK
redis> SADD some-string bar
(error) WRONGTYPE Operation against a key
 holding the wrong kind of value
\end{minted}

@@

\begin{minted}[xleftmargin=1em]{text}
redis> SET some-string foo
OK
redis> INCR some-string
(error) ERR value is not an integer or out
 of range
\end{minted}

\paragraph{In Hedis} The same goes for Hedis, since it's only a simple wrapping
 on top of Redis's protocol written in Haskell.

\begin{minted}[xleftmargin=1em]{haskell}
program :: Redis (Either Reply Integer)
program = do
    set "some-string" "foo"
    sadd "some-string" ["a"]
\end{minted}

It yields the same error as in Redis's client.

\begin{minted}[xleftmargin=1em]{haskell}
Left (Error "WRONGTYPE Operation against a
 key holding the wrong kind of value")
\end{minted}

\paragraph{The Cause} These problems arise from the absence of type checking,
with respects to \textbf{the type of a value that associates with a key}.

@@

Haskell makes it easy to build and use Domain Specific Languages (DSLs),
and Hedis can be regarded as one of them. What makes Hedis special is that,
it has variable bindings (between keys and values), but with very
little or no semantic checking, neither dynamically nor statically.

We began with making Hedis a dynamically checked embedded DSL, and implemented a
runtime type checker that keeps track of everything. But then we found that
things can be a lot easier, by leveraging the host language's type checker,
which also make it statically type-checked!

@@

To summarize our contributions:

\begin{itemize}[noitemsep]
\item We make Hedis statically type-checked, without runtime overhead.
\item We demonstrates how to model variable bindings of an embedded DSL with
language extensions like type-level literals and data kinds.
\item We provide (yet another) an example of encoding effects and
 constraints of an action in types, with indexed monad\cite{indexedmonad}
  and other language extensions such as closed
  type-families\cite{closedtypefamilies} and constraints
  kinds\cite{constraintskinds}.
\item Edis, a package we built for programmers. This package helps programmers
 to write more reliable Redis codes, and also makes Redis polymorphic by
 automatically converting back and forth from values of any types and boring
 ByteStrings.
\end{itemize}

\section{Type-level Dictionary}

To check the bindings between keys and values, we need a \emph{dictionary-like}
structure, and encode it as a \emph{type} somehow.

\subsection{Datatype promotion}
Normally, at the term level, we could express data type of the dictionary with
\emph{type synonym} like this.

\begin{minted}[xleftmargin=1em]{haskell}
type Key = String
type Dictionary = [(Key, TypeRep)]
\end{minted}

To encode this in the type level, everything has to be
\emph{promoted}\cite{promotion} one level up.
From terms to types, and from types to kinds.

Luckily, with recently added GHC extension \emph{data kinds}. With data kinds
we can define our own kinds just like we can with datatypes, and every suitable
datatype will be promoted to be a kind, and its value constructors promoted
to be type constructors, automatically.

\begin{minted}[xleftmargin=1em]{haskell}
data List a = Nil | Cons a (List a)
\end{minted}

Give rise to the following kinds and type constructors:\footnotemark\footnotemark

\footnotetext{To distinguish between types and promoted constructors that have
 ambiguous names, prefix promoted constructor with a single quote like
 \mintinline{text}{'Nil} and \mintinline{text}{'Cons}}
\footnotetext{All kinds have \emph{sort} BOX in Haskell\cite{sorts}}

\begin{minted}[xleftmargin=1em]{haskell}
List k :: BOX
Nil  :: List k
Cons :: k -> List k -> List k
\end{minted}

Haskell sugars lists \mintinline{haskell}{[1, 2, 3]} and
tuples \mintinline{text}{(1, 'a')} with brackets and parentheses.
Promoted lists and tuples are also sugared with single quote prefixed. For
example \mintinline{text}{'}\mintinline{haskell}{[Int, Char]},
\mintinline{text}{'}\mintinline{haskell}{(Int, Char)}.

\subsection{Type-level literals}

Now we have type-level lists and tuples to construct the dictionary.
For keys, \mintinline{haskell}{String} also has a type-level correspondence:
\mintinline{haskell}{Symbol}.

\begin{minted}[xleftmargin=1em]{haskell}
data Symbol
\end{minted}

Symbol is defined without a value constructor, because it's
intended to be used as a promoted kind.

\begin{minted}[xleftmargin=1em]{haskell}
"this is a type-level string literal" :: Symbol
\end{minted}

Nonetheless, it's still useful to have a term-level value that links with a
Symbol, when we want to retreive type-level informations at runtime (but not the
 other way around!).

\subsection{Putting everything together}

With all of these ingrediants ready, let's build some dictionaries!

\begin{minted}[xleftmargin=1em]{haskell}
type DictEmpty = '[]
type Dict0 = '[ '("key", Bool) ]
type Dict1 = '[ '("A", Int), '("B", "A") ]
\end{minted}

They are defined with \emph{type synonym}, since they are \emph{types}
, not \emph{terms}. If we ask \mintinline{text}{GHCi} what is the kind of
\mintinline{haskell}{Dict1}, we get \mintinline{haskell}{Dict1 :: [ (Symbol, *) ]}

The star \mintinline{haskell}{*} is a kind, it stands for any kind of types,
such as \mintinline{haskell}{Int}, \mintinline{haskell}{Char} or even a symbol
\mintinline{haskell}{"symbol"}, while \mintinline{haskell}{Symbol} stands only
for all symbols.

\section{Indexed Monad}

Redis commands are \emph{actions}.
We could capture the effects caused by an action, by expressing the states it
affects, before and after. That is, the \emph{preconditions} and
\emph{postconditions} of an action. In such way, we could also impose
constraints on the preconditions.

\emph{Indexed monads} (or \emph{monadish},
\emph{parameterised monad})\cite{indexedmonad}
can be used\cite{typefun}\cite{staticresources} to model such preconditions and
postconditions in types. An indexed monad is a type constructor that takes three
arguments: an intitial state, a final state, and the type of a value that an
action computes, which can be read like a Hoare triple\cite{kleisli}.

\begin{minted}[xleftmargin=1em]{haskell}
class IMonad m where
    unit :: a -> m p p a
    bind :: m p q a -> (a -> m q r b) -> m p r b
\end{minted}

Class \mintinline{haskell}{IMonad} comes with two operations:
\mintinline{haskell}{unit} for identities and \mintinline{haskell}{bind} for
compositions, as in Monad.

We define a new datatype \mintinline{haskell}{Edis}, which is basically just
\mintinline{haskell}{Redis} indexed with more informations in types. And we make
 it an instance of \mintinline{haskell}{IMonad}.

\begin{minted}[xleftmargin=1em]{haskell}
newtype Edis p q a = Edis { unEdis :: Redis a }

instance IMonad Edis where
    unit = Edis . return
    bind m f = Edis (unEdis m >>= unEdis . f )
\end{minted}

The first and second argument of type \mintinline{haskell}{Edis} is where the
dictionaries going to stay.

\subsection{The first attempt}

In Redis, \mintinline{text}{PING} does nothing but replies with
\mintinline{text}{PONG} if the connection is alive. In Hedis,
\mintinline{text}{PING} has type:

\begin{minted}[xleftmargin=1em]{haskell}
ping :: Redis (Either Reply Status)
\end{minted}

Now we have \mintinline{haskell}{Edis}, let's make our own version of
\mintinline{haskell}{ping}\footnotemark

\begin{minted}[xleftmargin=1em]{haskell}
ping :: Edis xs xs (Either Reply Status)
ping = Edis Hedis.ping
\end{minted}

Dicionary \mintinline{haskell}{xs} in the type remain unaffected after the
action, because \mintinline{haskell}{ping} does not affect any key-value
bindings. To encode other commands that alters key-value bindings, we need
 type-level functions to annotate those effects on the dictionary.

\section{Bookkeeping with type families}
\subsection{Closed type families}

Type families have a wide variaty of applications. They can appear inside type
classes\cite{tfclass}\cite{tfsynonym}, or at toplevel. Toplevel type families
can be used to compute over types, they come in open\cite{tfopen} or closed
forms\cite{tfclosed}. We are using \emph{closed type families} since we need
none of the extensibility provided by open type families, and we may need to
overlap instances, like what we do in defining function cases.

%  are
% permitted as well.
%
%
%  open type synonym families, and
% closed type synonym families
% One of which is to compute types, we are using closed type families, since all
% we want is to use it as a function on the type level, no extensibility needed,
% everything is defined in one-piece

\footnotetext{\mintinline{haskell}{ping} from Hedis is qualified with
\mintinline{haskell}{Hedis} to prevent function name clashing}
%
% \mintinline{haskell}{IMonad} \mintinline{haskell}{IMonad} takes three arguments: an intitial state \mintinline{haskell}{p}, a final
%  state, and computes a value of type a

\subsection{Proxies and Kind polymorphism}

Since Haskell has a phase distinction\cite{phasedistinction} between

To link a symbol with a term-level value

What we need is a \emph{singleton}\cite{singletons}, a type
that has only one instance at the term level.

With \emph{kind polymorphism}\cite{promotion}, we have parametric polymorphism
at the kind level.

\begin{minted}[xleftmargin=1em]{haskell}
Proxy :: Proxy "a"
\end{minted}

\subsection{Closed type families}

\subsection{Conclusions and Related Work}

% references
\bibliographystyle{abbrvnat}
\bibliography{cites}

% \subsection{Model bindings with Type-level Dictionaries}
% \subsection{Bookkeeping with Type families}
% \subsection{Capture effects with Index Monad}
% \paragraph{Type-level literals}
